### 2014-04-30 ApTranscriptome_TR.Rmd
Updated script to use DeconSeq "cleaned" transcriptome assembly that removed about 5k bacterial, archael, viral and human contigs.
Surprisingly, after re-running Sailfish, this increased the number of responsive
transcripts, after adjusting P-values, to 30k!!!

# Load Bioconductor libraries
#source("http://bioconductor.org/biocLite.R")
#biocLite("topGO")
#biocLite("Rgraphviz")
library(topGO) 
#suppressMessages(library(Rgraphviz))

@@

Two ant colonies were used for the transcriptome sequencing. The first, designated A22, was collected at Molly Bog, Vermont in August 2012 by Nick Gotelli and Andrew Nguyen. This colony was putatively identifed as *A. picea*. The second colony, designated Ar, was collected by Lauren Nichols in Raleigh, North Carolina. These colonies were maintained in the lab for 6 months prior to sample collection.

For each colony, three ants were exposed to one of 12 temperature treatments, every 3.5C ranging from 0C to 38.5C, for one hour in glass tubes in a water bath. The ants were flash frozen and stored at -80C until RNA was extracted using a two step extraction; [RNAzol RT](http://www.mrcgene.com/rnazol.htm) (Molecular Research Center, Inc) followed by an [RNeasy Micro](http://www.qiagen.com/products/catalog/sample-technologies/rna-sample-technologies/total-rna/rneasy-micro-kit) column (Qiagen). Samples from each colony were pooled and sequenced in separate lanes on a 100bp paired-end run of an Illumina HiSeq at the University of Minnesota Genomics Center, yielding 20e6 and 16e6 reads for the A22 and Ar samples, respectively.

@@

# extract significant transcripts
sig.TPM.dt.sub <- TPM.dt.sub[signif.transcripts$Transcript]
```

@@

png(file = paste(resultsdir, "responsive_transcripts_mosaicplot.png", sep = ""))
mosaicplot(tt2, shade = TRUE, xlab = "ApVT", ylab = "AcNC", main = "Mosaic plot of responsive transcripts")
dev.off()

@@

Yes - transcripts that are up-regulated at high temperatures in *A22* are more highly-expressed at 19.25C in *Ar*. Note that A22 had the larger library size so if this was due to TPM not correctly accounting for differences in reads between samples, we would expect to see a positive instead of negative value here.

@@

@@

The 'Intermediate' expressed transcripts are core molecular processes that are expressed at non-stressful temperatures, and shut-off when the organism experiences thermall stress. We hypothesized that if the more southern *Ar* colony was more thermally-tolerant than *A22*, transcripts with 'Intermediate' expression (10-30C) would be active across a wider range of temperatures. To test this with our data, we calculated the standard deviation of the expression function for each temperature transcript that was 'Intermediate' expressed in each colony.

@@

@@

# t-test to compare standard deviation of 'Bimodal' transcripts between colonies
t2 <- t.test(Ar.bim.sd, A22.bim.sd)
```

@@

# t-test
t3 <- t.test(varshape.out$s.offset, alternative = "two.sided")

@@

# t-test
t4 <- t.test(varshape.out$s.slope, alternative = "two.sided")

# curvature = curvature1 - curvature2
varshape.out$curvature <- varshape.out$A22.curve - varshape.out$Ar.curve
# to compare among transcripts that differ in scale of expression, standardize by overall mean for each transcript
varshape.out$s.curvature <- NA
for(i in 1:nrow(varshape.out)) {
  varshape.out[i, "s.curvature"] <- (varshape.out[i, "curvature"] / mean(c(varshape.out[i, "A22.curve"], varshape.out[i, "Ar.curve"])))
}
# change 'Inf' values to 0
varshape.out[which(varshape.out$Ar.curve == 0 & varshape.out$A22.curve == 0), "s.curvature"] <- 0
# t-test
tt.test(varshape.out$s.curvature, alternative = "two.sided")

@@

# t-test
t.test(varshape.out$s.wiggle, alternative = "two.sided")
```

>>>>

# Load Bioconductor libraries
#source("http://bioconductor.org/biocLite.R")
#biocLite("topGO")
#biocLite("Rgraphviz") # INSTALL BROKEN IN R 3.1.0
library(topGO) 
#suppressMessages(library(Rgraphviz))

@@

Two ant colonies were used for the transcriptome sequencing. The first, designated *A22*, was collected at Molly Bog, Vermont in August 2012 by Nick Gotelli and Andrew Nguyen. The second colony, designated *Ar*, was collected by Lauren Nichols in Raleigh, North Carolina. These colonies were identified as *A. picea* and *A. carolinensis* by Bernice Bacon de Marco. The colonies were maintained in the lab for 6 months prior to sample collection.

For each colony, three ants were exposed to one of 12 temperature treatments, every 3.5C ranging from 0C to 38.5C, for one hour in glass tubes in a water bath. The ants were flash frozen and stored at -80C until RNA was extracted using a two step extraction; [RNAzol RT](http://www.mrcgene.com/rnazol.htm) (Molecular Research Center, Inc) followed by an [RNeasy Micro](http://www.qiagen.com/products/catalog/sample-technologies/rna-sample-technologies/total-rna/rneasy-micro-kit) column (Qiagen). Samples from each colony were pooled and sequenced in separate lanes on a 100bp paired-end run of an Illumina HiSeq at the University of Minnesota Genomics Center, yielding 20^6^ and 16^6^ reads for the *A22* and *Ar* samples, respectively.

@@

# extract significant transcripts
setkey(TPM.dt.sub, Transcript)
sig.TPM.dt.sub <- TPM.dt.sub[signif.transcripts$Transcript]
```

@@

png(file = paste(resultsdir, "responsive_transcripts_mosaicplot.png", sep = ""), width = 600, height = 480)
mosaicplot(tt2, xlab = "ApVT", ylab = "AcNC", cex = 0.7, las = 1, main = "")
dev.off()

@@

Yes - transcripts that are up-regulated at high temperatures in *A22* are more highly-expressed at 19.25°C in *Ar*. Note that A22 had the larger library size so if this was due to TPM not correctly accounting for differences in reads between samples, we would expect to see a positive instead of negative value here.

@@

The overall pattern that emerges from these 4 tests is that constitutive expression at optimal conditions of 19.5°C is overall greater for *Ar* than *A22*.

@@

The 'Intermediate' expressed transcripts are core molecular processes that are expressed at non-stressful temperatures, and shut-off when the organism experiences thermall stress. We hypothesized that if the more southern *Ar* colony was more thermally-tolerant than *A22*, transcripts with 'Intermediate' expression (10-30°C) would be active across a wider range of temperatures. To test this with our data, we calculated the standard deviation of the expression function for each temperature transcript that was 'Intermediate' expressed in each colony.

@@

```{r plot_thermal_breadth2, echo = FALSE, eval = TRUE}
# predict expression for these transcripts
int.predicted <- matrix(nrow = 0, ncol = 5)

# for A22
colony = "A22"

for(i in 1:length(A22.int.lm)) {
    lmitem <- A22.int.lm[[i]]
    pred.vals <- seq(from = 0, to = 38.5, by = 2.5)
    newdf <- data.frame(Transcript = names(A22.int.lm[i]), colony = rep(colony, length = length(pred.vals)), val = pred.vals)
    pout <- predict(lmitem, newdata=newdf)
    pout <- cbind(newdf, logTPM = pout) # note this is on log scale
    # back-transform from log-scale
    pout$pTPM <- exp(pout$logTPM)
    # change any negative values to zero
    pout$pTPM <- ifelse(pout$pTPM < 0, 0, round(pout$pTPM, 5))
    # concatenate
    int.predicted <- rbind(int.predicted, pout)
}

# for Ar
colony = "Ar"

for(i in 1:length(Ar.int.lm)) {
    lmitem <- Ar.int.lm[[i]]
    pred.vals <- seq(from = 0, to = 38.5, by = 2.5)
    newdf <- data.frame(Transcript = names(Ar.int.lm[i]), colony = rep(colony, length = length(pred.vals)), val = pred.vals)
    pout <- predict(lmitem, newdata=newdf)
    pout <- cbind(newdf, logTPM = pout) # note this is on log scale
    # back-transform from log-scale
    pout$pTPM <- exp(pout$logTPM)
    # change any negative values to zero
    pout$pTPM <- ifelse(pout$pTPM < 0, 0, round(pout$pTPM, 5))
    # concatenate
    int.predicted <- rbind(int.predicted, pout)
}

# change to data.table
int.predicted <- data.table(int.predicted)
# scale transcripts so can compare
int.predicted[,pTPM.scaled:=scale(pTPM), by = Transcript]
str(int.predicted)

# plot
intplot <- ggplot(int.predicted, aes(x=val, y=pTPM.scaled, group=Transcript)) + 
  geom_line() + 
  facet_grid(. ~ colony)
intplot
```

Plot shows all the intermediate transcripts, though not possible to tell that *Ar* transcripts have greater variance than *A22* transcripts.

@@

# t-test to compare standard deviation of 'Bimodal' transcripts between colonies
(t2 <- t.test(Ar.bim.sd, A22.bim.sd))
```

@@

# t-test
(t3 <- t.test(varshape.out$s.offset, alternative = "two.sided"))

@@

# t-test
(t4 <- t.test(varshape.out$s.slope, alternative = "two.sided"))

# curvature = curvature1 - curvature2
varshape.out$curvature <- varshape.out$A22.curve - varshape.out$Ar.curve
# to compare among transcripts that differ in scale of expression, standardize by overall mean for each transcript
varshape.out$s.curvature <- NA
for(i in 1:nrow(varshape.out)) {
  varshape.out[i, "s.curvature"] <- (varshape.out[i, "curvature"] / mean(c(varshape.out[i, "A22.curve"], varshape.out[i, "Ar.curve"])))
}
# change 'Inf' values to 0
varshape.out[which(varshape.out$Ar.curve == 0 & varshape.out$A22.curve == 0), "s.curvature"] <- 0
# t-test
(t5 <- t.test(varshape.out$s.curvature, alternative = "two.sided"))

@@

# t-test
(t6 <- t.test(varshape.out$s.wiggle, alternative = "two.sided"))
```
